Exact Phylodynamics via Structured Markov Genealogy Processes

### 2024-05-21 ms.Rnw
new phylopomp version

<<prefix,include=FALSE,cache=FALSE,purl=FALSE>>=
prefix <- "sgp"
source("setup.R")
@
<<packages,include=FALSE,cache=FALSE>>=
library(tidyverse)
library(ggtree)
library(pomp)
library(cowplot)
library(viridis)
library(phylopomp)
stopifnot(getRversion() >= "4.3")
stopifnot(packageVersion("pomp")>="5.5")
stopifnot(packageVersion("phylopomp")>="0.10.3")
theme_set(theme_bw(base_family="serif"))
options(
  width=150,
  keep.source=TRUE,
  encoding="UTF-8",
  stringsAsFactors=FALSE,
  dplyr.summarise.inform=FALSE,
  pomp_archive_dir="results"
)
set.seed(1159254136)
@

@@

\begin{table}[h!]
  \caption{
    \label{tab:sirs_model_elements}
    Elements of the SIRS model pertinent to the genealogy process.
  }
  \renewcommand{\arraystretch}{1.2}
  \begin{tabular}{cclc}
    \hline\hline
    $u$ & $\alpha_u$ & Event type & $r_u$ \\
    \hline
    $\lab{Trans}$  & ${\beta S I}/{N}$ & pure birth  & 2 \\
    $\lab{Recov}$  & $\gamma I$        & pure death  & 0 \\
    $\lab{Wane}$   & $\omega R$        & neutral     & 0 \\
    $\lab{Sample}$ & $\psi I$          & pure sample & 1 \\
    \hline\hline
  \end{tabular}
\end{table}

@@

\begin{table}[h!]
  \caption{
    \label{tab:seirs_model_elements}
    Elements of the SEIRS model pertinent to the genealogy process.
  }
  \renewcommand{\arraystretch}{1.2}
  \begin{tabular}{cclc}
    \hline\hline
    $u$ & $\alpha_u$ & Event type & $r_u$ \\
    \hline
    $\lab{Trans}$  & ${\beta S I}/{N}$ & pure birth     & $(1,1)$ \\
    $\lab{Prog}$   & $\sigma E$        & pure migration & $(0,1)$ \\
    $\lab{Recov}$  & $\gamma I$        & pure death     & $(0,0)$ \\
    $\lab{Wane}$   & $\omega R$        & neutral        & $(0,0)$ \\
    $\lab{Sample}$ & $\psi I$          & pure sample    & $(0,1)$ \\
    \hline\hline
  \end{tabular}
\end{table}

@@

params |>
  with(
    mcap(logLik,sigma,span=0.25)
  ) -> mcap
@

\begin{figure}
  \begin{center}
    <<seirs3_plot,fig.dim=c(8,2.8),out.width="100%">>=
    plot_grid(
      A=seirs_tree |>
        plot(points=TRUE,palette="#000000")+
        labs(x="time"),
      B=params |>
        ggplot()+
        geom_point(aes(x=sigma,y=logLik))+
        geom_line(data=mcap$fit,aes(x=parameter,y=smoothed),color="blue")+
        geom_vline(xintercept=seirs.params$sigma,color="red")+
        geom_vline(xintercept=mcap$ci,linetype=2)+
        geom_hline(
          yintercept=max(mcap$fit$smoothed)-c(0,mcap$delta),
          linetype=2
        )+
        ##        lims(y=c(max(params$logLik)-12,NA))+
        labs(
          color=character(0),
          y="log likelihood",
          x=expression(sigma)
        )+
        theme_classic(),
      labels="AUTO",
      nrow=1,
      rel_widths=c(3,4)
    )
    @
  \end{center}
  \caption{
    \label{fig:seirs_example}
    Likelihood computation for the SEIRS model by sequential Monte Carlo.
    (A)~Simulated genealogy for $\beta=\Sexpr{seirs.params$Beta}$,
    $\sigma=\Sexpr{seirs.params$sigma}$,
    $\gamma=\Sexpr{seirs.params$gamma}$,
    $\psi=\Sexpr{seirs.params$psi}$,
    $\omega=\Sexpr{seirs.params$omega}$,
    $(S_0,E_0,I_0,R_0)=(\Sexpr{with(seirs.params,c(S0,E0,I0,R0))})$.
    (B)~Likelihood slice in the $\sigma$-direction.
    Each point represents the estimate of a distinct sequential Monte Carlo computation.
    The blue curve shows a LOESS smooth;
    the dashed vertical lines enclose the Monte Carlo-adjusted 95\% confidence interval \citep{Ionides2017}.
  }
\end{figure}

@@

\begin{table}[h!]
  \caption{
    \label{tab:lbdp_model_elements}
    Elements of the linear birth-death-sampling model pertinent to the genealogy process.
  }
  \renewcommand{\arraystretch}{1.2}
  \begin{tabular}{cclc}
    \hline\hline
    $u$ & $\alpha_u$ & Event type & $r_u$ \\
    \hline
    $\lab{Birth}$  & $\lambda N$  & pure birth  & 2 \\
    $\lab{Death}$  & $\gamma N$   & pure death  & 0 \\
    $\lab{Sample}$ & $\psi N$     & pure sample & 1 \\
    \hline\hline
  \end{tabular}
\end{table}

>>>>

<<prefix,include=FALSE,cache=FALSE,purl=FALSE>>=
prefix <- "sgp"
source("setup.R")
@
<<packages,include=FALSE,cache=FALSE>>=
library(tidyverse)
library(ggtree)
library(pomp)
library(cowplot)
library(viridis)
library(phylopomp)
stopifnot(getRversion() >= "4.3")
stopifnot(packageVersion("pomp")>="5.8")
stopifnot(packageVersion("phylopomp")>="0.11")
theme_set(theme_bw(base_family="serif"))
options(
  width=150,
  keep.source=TRUE,
  encoding="UTF-8",
  stringsAsFactors=FALSE,
  dplyr.summarise.inform=FALSE,
  pomp_archive_dir="results"
)
set.seed(1159254136)
@

@@

\begin{table}[h!]
  \caption{
    \label{tab:sirs_model_elements}
    Elements of the SIRS model pertinent to the genealogy process.
  }
  %% \renewcommand{\arraystretch}{1.2}
  \begin{tabular}{cccl}
    \hline\hline
    $u$ & $\alpha_u$ & $r_u$ & Event type \\
    \hline
    $\lab{Trans}$  & $\frac{\beta S I}{N}$ \bigstrut & 2 & pure birth \\
    $\lab{Recov}$  & $\gamma I$ \bigstrut & 0 & pure death \\
    $\lab{Wane}$   & $\omega R$ \bigstrut & 0 & neutral \\
    $\lab{Sample}$ & $\psi I$ \bigstrut & 1 & pure sample \\
    \hline\hline
  \end{tabular}
\end{table}

@@

\input{figs/seirs_events}

\begin{table}[h!]
  \caption{
    \label{tab:seirs_model_elements}
    Elements of the SEIRS model pertinent to the genealogy process.
  }
  %% \renewcommand{\arraystretch}{1.2}
  \begin{tabular}{ccccl}
    \hline\hline
    $u$ & $\alpha_u$ & $x\mapsto{x'}$ & $r_u$ & Event type \\
    \hline
    $\lab{Trans}$  & $\frac{\beta S I}{N}$ \bigstrut & $(S,E)\mapsto(S-1,E+1)$  \bigstrut & $(1,1)$ & pure birth \\
    $\lab{Prog}$   & $\sigma E$ \bigstrut & $(E,I)\mapsto(E-1,I+1)$ \bigstrut & $(0,1)$ & pure migration \\
    $\lab{Recov}$  & $\gamma I$ \bigstrut & $(I,R)\mapsto(I-1,R+1)$ \bigstrut & $(0,0)$ & pure death \\
    $\lab{Wane}$   & $\omega R$ \bigstrut & $(S,R)\mapsto(S+1,R-1)$ \bigstrut & $(0,0)$ & neutral \\
    $\lab{Sample}$ & $\psi I$ \bigstrut & $x\mapsto{x}$ \bigstrut & $(0,1)$ & pure sample \\
    \hline\hline
  \end{tabular}
\end{table}

\begin{table}[h!]
  \caption{
    \label{tab:seirs_integ}
    Elements of a scheme for computing the likelihood under the SEIRS model.
    The $y\mapsto{y'}$ column depicts the proposed painting schematically.
    Blue is used for the $\lab{E}$ deme and yellow for the $\lab{I}$ deme.
  }
  %% \renewcommand{\arraystretch}{1.2}
  \begin{tabular}{c c c c c c c c | c}
    \hline\hline
    & $u$ & $s$ & $\BinRatio{n}{\ell}{r}{s}$ & $y\mapsto{y'}$ & $\pi$ & n & $Q$ & Line\\[2ex]
    \hline
    \multirow[c]{10}{*}[-1ex]{\rotatebox{90}{$t\notin{{B}\cup{S_0}\cup{S_1}}$}}
    & \multirow{4}{*}{$\lab{Trans}$}
    & $(0,0)$ & $\left(\frac{E-\ell_E}{E}\right)\left(\frac{I-\ell_I}{I}\right)$ \bigstrut & \unbroken \bigstrut & $\frac{I+1-\ell_I}{I+1}$ & 1 & 1 & 1 \\
    \cline{3-8}
    && $(1,0)$ & $\frac{I-\ell_I}{E I}$ \bigstrut & \maizeblue \bigstrut & $\frac{1}{2(I+1)}$ & $\ell_I$ & 1 & 2 \\
    \cline{3-8}
    && $(0,1)$ & $\frac{E-\ell_E}{E I}$ \bigstrut & \maizemaize \bigstrut & $\frac{1}{2(I+1)}$ & $\ell_I$ & 1 & 3 \\
    \cline{3-8}
    && $(1,1)$ & --- & --- & 0 & 0 & 0 & 4 \\
    \cline{2-8}
    &\multirow{2}{*}{$\lab{Prog}$}
    & $(0,0)$ & $\frac{I-\ell_I}{I}\,\Indicator{E\ge\ell_E}$ \bigstrut & \unbroken \bigstrut & $\frac{E+1-\ell_E}{E+1}\,\Indicator{E>\ell_E}$ \bigstrut & 1 & 1 & 5 \\
    \cline{3-8}
    && $(0,1)$ & $\frac{1}{I}\,\Indicator{E\ge\ell_E}$ \bigstrut & \bluemaize \bigstrut & $\frac{1}{E+1}\,\Indicator{E>\ell_E}$ \bigstrut & $\ell_E$ & 1 & 6 \\
    \cline{2-8}
    & $\lab{Recov}$ & $(0,0)$ & $\Indicator{I\ge\ell_I}$ \bigstrut & \unbroken \bigstrut & $\Indicator{I>\ell_I}$ \bigstrut & 1 & 1 & 7 \\
    \cline{2-8}
    & $\lab{Wane}$ & $(0,0)$ & 1 \bigstrut & \unbroken \bigstrut & 1 & 1 & 1 & 8 \\
    \cline{2-8}
    & $\lab{Sample}$ & $\ast$ & --- & --- & 1 & 0 & 0 & 9 \\
    \cline{1-8}
    \multirow[c]{4}{*}[-3ex]{\rotatebox{90}{$t\in{B}$}}
    & \multirow{4}{*}{$\lab{Trans}$}
    & \multirow{4}{*}{$(1,1)$}
    & $\frac{1}{E I}$ \bigstrut & \branchup \bigstrut & $\frac{1}{2}$ \bigstrut & 1 & 1 & 10 \\
    \cline{4-8}
    &&& $\frac{1}{E I}$ \bigstrut & \branchdown \bigstrut & $\frac{1}{2}$ \bigstrut & 1 & 1 & 11 \\
    \cline{4-8}
    &&& --- \bigstrut & \branchupnix \bigstrut & 0 & 0 & 0 & 12 \\
    \cline{4-8}
    &&& --- \bigstrut & \branchdownnix \bigstrut & 0 & 0 & 0 & 13 \\
    \cline{2-8}
    & $\ast$ & $\ast$ & --- \bigstrut & --- & 0 & 0 & 0 & 14 \\
    \cline{1-8}
    \multirow[c]{2}{*}[-2ex]{\rotatebox{90}{$t\in{S_0}$} \bigstrut}
    & \multirow{2}{*}{$\lab{Sample}$}
    & \multirow{2}{*}{$(0,0)$}
    & $\frac{I-\ell_I}{I}$ \bigstrut & \samplezero & 1 & 1 & 1 & 15 \\
    \cline{4-8}
    &&& --- & \samplezeronix \bigstrut & 0 & 0 & 0 & 16 \\
    \cline{2-8}
    & $\ast$ & $\ast$ & --- & --- & 0 & 0 & 0 & 17 \\
    \cline{1-8}
    \multirow[c]{2}{*}[-2ex]{\rotatebox{90}{$t\in{S_1}$}}
    & \multirow{2}{*}{$\lab{Sample}$}
    & \multirow{2}{*}{$(0,1)$}
    & $\frac{1}{I}$ \bigstrut & \sampleone \bigstrut & 1 & 1 & 1 & 18 \\
    \cline{4-8}
    &&& --- & \sampleonenix \bigstrut & 0 & 0 & 0 & 19 \\
    \cline{2-8}
    & $\ast$ & $\ast$ & --- & --- & 0 & 0 & 0 & 20 \\
    \hline\hline
  \end{tabular}
\end{table}

A variety of importance-sampling kernels are permissible under the terms of \cref{thm:obsc_uncond}.
Given a particular choice of importance-sampling kernel, the filtering equation follows.
For example, with the choices as in \cref{tab:seirs_integ}, the filtering equation is as follows.
As previously, let $B$ be the set of branch times, $S_0$ be the set of tip-sample times, and $S_1$ be the set of inline sample times.
Then for $t\notin{{B}\cup{S_0}\cup{S_1}}$, we have
\begin{mathsize}{8pt}{9pt}
  \begin{equation*}
    \begin{split}
      \frac{\partial{w}}{\partial{t}}(t,S,E,I,R,y)=
      %% &\frac{\beta\,(S+1)\,I}{N}\,\frac{I-\ell_I}{I}\,\frac{E-\ell_E}{E}\,w(t,S+1,E-1,I,R,y)\\
      %% &+\sum_{k=1}^{\ell_I}\frac{\beta\,(S+1)\,I}{N}\,\frac{1}{I}\,\frac{I-\ell_I}{E}\,w(t,S+1,E-1,I,R,\leftlim{y}_2)\\
      %% &+\sum_{k=1}^{\ell_I}\frac{\beta\,(S+1)\,I}{N}\,\frac{1}{I}\,\frac{E-\ell_E}{E}\,w(t,S+1,E-1,I,R,\leftlim{y}_3)
      %% -\frac{\beta\,S\,I}{N}\,w(t,S,E,I,R,y)\\
      %% &+\sigma\,(E+1)\frac{E+1-\ell_E}{E+1}\,\Indicator{E\ge{\ell_E}}\,\frac{I-\ell_I}{I}\,\frac{E+1}{E+1-\ell_E}\,w(t,S,E+1,I-1,R,y)\\
      %% &+\sum_{k=1}^{\ell_E}\sigma\,(E+1)\frac{1}{E+1}\,\Indicator{E\ge{\ell_E}}\,\frac{E+1}{I}\,w(t,S,E+1,I-1,R,\leftlim{y}_6)
      %% -\sigma\,E\,w(t,S,E,I,R,y)\\
      %% &+\gamma\,(I+1)\,\Indicator{I\ge\ell_I}\,w(t,S,E,I+1,R-1,y)
      %% -\gamma\,I\,w(t,S,E,I,R,y)\\
      %% &+\omega\,(R+1)\,w(t,S-1,E,I,R+1,y)
      %% -\omega\,R\,w(t,S,E,I,R,y)
      %% -\psi\,I\,w(t,S,E,I,R,y).
      %%
      &\frac{\beta\,(S+1)\,I}{N}\,\left(1-\frac{\ell_I}{I}\right)\,\left(1-\frac{\ell_E}{E}\right)\,w(t,S+1,E-1,I,R,y)\\
      &+\sum_{k=1}^{\ell_I}\frac{\beta\,(S+1)\,I}{N}\,\,\frac{1}{E}\,\left(1-\frac{\ell_I}{I}\right)\,w(t,S+1,E-1,I,R,\leftlim{y}_2^k)\\
      &+\sum_{k=1}^{\ell_I}\frac{\beta\,(S+1)\,I}{N}\,\frac{1}{I}\,\left(1-\frac{\ell_E}{E}\right)\,w(t,S+1,E-1,I,R,\leftlim{y}_3^k)
      -\frac{\beta\,S\,I}{N}\,w(t,S,E,I,R,y)\\
      &+\sigma\,(E+1)\,\Indicator{E\ge{\ell_E}}\,\left(1-\frac{\ell_I}{I}\right)\,w(t,S,E+1,I-1,R,y)\\
      &+\sum_{k=1}^{\ell_E}\sigma\,(E+1)\,\Indicator{E\ge{\ell_E}}\,\frac{1}{I}\,w(t,S,E+1,I-1,R,\leftlim{y}_6^k)
      -\sigma\,E\,w(t,S,E,I,R,y)\\
      &+\gamma\,(I+1)\,\Indicator{I\ge\ell_I}\,w(t,S,E,I+1,R-1,y)
      -\gamma\,I\,w(t,S,E,I,R,y)\\
      &+\omega\,(R+1)\,w(t,S-1,E,I,R+1,y)
      -\omega\,R\,w(t,S,E,I,R,y)
      -\psi\,I\,w(t,S,E,I,R,y).
      %%
      %% &\frac{\beta\,(S+1)\,I}{N}\,\left(1-\frac{\ell_E}{E}\right)\,w(t,S+1,E-1,I,R,y)\\
      %% &+\sum_{k=1}^{\ell_I}\frac{\beta\,(S+1)\,I}{N}\,\,\frac{1}{E}\,\left(1-\frac{\ell_I}{I}\right)\,w(t,S+1,E-1,I,R,\leftlim{y}_2^k)
      %% -\frac{\beta\,S\,I}{N}\,w(t,S,E,I,R,y)\\
      %% &+\sigma\,(E+1)\,\Indicator{E\ge{\ell_E}}\,\left(1-\frac{\ell_I}{I}\right)\,w(t,S,E+1,I-1,R,y)\\
      %% &+\sum_{k=1}^{\ell_E}\sigma\,(E+1)\,\Indicator{E\ge{\ell_E}}\,\frac{1}{I}\,w(t,S,E+1,I-1,R,\leftlim{y}_6^k)
      %% -\sigma\,E\,w(t,S,E,I,R,y)\\
      %% &+\gamma\,(I+1)\,\Indicator{I\ge\ell_I}\,w(t,S,E,I+1,R-1,y)
      %% -\gamma\,I\,w(t,S,E,I,R,y)\\
      %% &+\omega\,(R+1)\,w(t,S-1,E,I,R+1,y)
      %% -\omega\,R\,w(t,S,E,I,R,y)
      %% -\psi\,I\,w(t,S,E,I,R,y).
    \end{split}
  \end{equation*}
\end{mathsize}%

@@

params |>
  with(
    mcap(logLik,sigma,span=0.5)
  ) -> mcap
@

\begin{figure}
  \begin{center}
    <<seirs3_plot,fig.dim=c(8,2.8),out.width="100%",dependon="seirs3">>=
    plot_grid(
      A=seirs_tree |>
        plot(points=TRUE,palette="#000000")+
        labs(x="time"),
      B=params |>
        ggplot()+
        geom_point(aes(x=sigma,y=logLik))+
        geom_line(data=mcap$fit,aes(x=parameter,y=smoothed),color="blue")+
        geom_vline(xintercept=seirs.params$sigma,color="red")+
        geom_vline(xintercept=mcap$ci,linetype=2)+
        geom_hline(
          yintercept=max(mcap$fit$smoothed)-c(0,mcap$delta),
          linetype=2
        )+
        ##        lims(y=c(max(params$logLik)-12,NA))+
        labs(
          color=character(0),
          y="log likelihood",
          x=expression(sigma)
        )+
        theme_classic(),
      labels="AUTO",
      nrow=1,
      rel_widths=c(3,4)
    )
    @
  \end{center}
  \caption{
    \label{fig:seirs_example}
    Likelihood computation for the SEIRS model by sequential Monte Carlo.
    (A)~Simulated genealogy for $\beta=\Sexpr{seirs.params$Beta}$,
    $\sigma=\Sexpr{seirs.params$sigma}$,
    $\gamma=\Sexpr{seirs.params$gamma}$,
    $\psi=\Sexpr{seirs.params$psi}$,
    $\omega=\Sexpr{seirs.params$omega}$,
    $(S_0,E_0,I_0,R_0)=(\Sexpr{with(seirs.params,c(S0,E0,I0,R0))})$.
    (B)~Likelihood slice in the $\sigma$-direction.
    Each point represents the estimate of a distinct sequential Monte Carlo computation.
    The blue curve shows a LOESS smooth;
    the dashed vertical lines enclose the Monte Carlo-adjusted 95\% confidence interval \citep{Ionides2017}.
  }
\end{figure}

@@

\begin{table}[h!]
  \caption{
    \label{tab:lbdp_model_elements}
    Elements of the linear birth-death-sampling model pertinent to the genealogy process.
  }
  \renewcommand{\arraystretch}{1.2}
  \begin{tabular}{cccl}
    \hline\hline
    $u$ & $\alpha_u$ & $r_u$ & Event type \\
    \hline
    $\lab{Birth}$  & $\lambda N$ & 2 & pure birth \\
    $\lab{Death}$  & $\gamma N$ & 0 & pure death \\
    $\lab{Sample}$ & $\psi N$ & 1 & pure sample \\
    \hline\hline
  \end{tabular}
\end{table}
